One Model May Not Fit All: Subgroup Detection Using MOB

### 2023-10-20 _MOB_paper.Rnw
Comments about changes
Paper updates
Revised manuscript and analyses. Did not touch application to Rasch trees. Still quite some work to do, but this is as far as I could get.

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to a dataset from a study of \citep{Demi09}\todo{take that out?}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program. Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citep{Breetal:1984}. 
In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable. An example could be that children who are highly motivated and have good reading skills show higher average math exam grades. 
MOB generalizes this idea by identifying groups of persons, specified by (combinations of) covariates, that differ in a the parameters of a statistical model. An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.

The framework of MOB is very flexible and can be applied in combination with a variety of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS}. In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}. Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study. In application example 1 we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.

While for this first application example we assume that the administered test \todo{should we call a psychological test a test/an instrument/??? to distinguish it from the verb and noun for statistical test? (Mirka: I like the term instrument to distinguish it from a statistical test)} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample. Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders. If certain items pose an advantage or disadvantage to either group, they are said to violate measurement invariance or exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group. This can be done in a way that allows to detect DIF on top of any true group differences in ability. While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven 
way.

In the following section, we will give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}. An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}. Mixture modelling aims at identifying latent classes of persons with different properties. It can also be combined with covariates. A comparison of both approches is given by \citet{FriStrZei:2014}.

\section{The Algorithm Behind MOB}

What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?); Marjolein: please read and decide whether we do need a picture of a regression tree and/or mob tree here, if yes, simulate but with abstract x1 x2, no story}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}). These figures share the tree-structure in the upper part, but differ in their lower part. The starting point of the tree structure is at the very top of each tree. The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the 
bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.

The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates. For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups. The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups. The categorical covariate \texttt{Race}, which was coded in three categories in this example, has been divided into two groups of categories: Black and Hispanic vs.~White. For a binary covariate, \todo{like ... -- add name if is there one in the other example?} 
on the other hand, there is only one possible cutpoint, namely between the two categories. While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.

The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree. While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items. 
The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data. Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting. Variables are selected based on a statistical test for parameter differences, also termed a test for structural change. Structural change is present if, for example, the model parameters systematically differ for children from lower vs.~higher income families. At each node, the algorithm will select the covariate showing the strongest structural change as the next splitting variable. Within this variables, the optimal cutpoint is chosen in a separate step.

The way the variable selection and cutpoint selection steps are separated in modern classification and regression tree and MOB algorithms, including the ones used here, distinguish them from the traditional classification and regression tree algorithms of \citet{Breetal:1984} and \citet{Qui:1993}. The traditional algorithms performed variable and cutpoint selection in one step. However, this leads to an undesirable behavior called variable selection bias. It means that the traditional algorithms prefer variables offering more cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. Modern algorithms for classification and regression trees that have solved this problem are the unbiased approaches \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, available in R in the \texttt{party} and \texttt{partykit} packages \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper.

Once there are no more covariates that show a significant structural change in any node, the MOB algorithm stops splitting. In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection. Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected. This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and partly also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected. Other stopping criteria are based on the number of persons in the end nodes. These crieria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}

In summary, the rationale of a MOB algorithm consists of the following steps:

\todo{why are the spaces so big? I didn't change the enumerate command}
\begin{enumerate}
\item The model parameters are first estimated jointly for all persons in the current node, starting with the full sample.
\item Structural change in the model parameters is assessed with respect to each available covariate.
\item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
\item Steps 1--3 are repeated recursively in the resulting groups until there is no more significant structural change (or the groups becomes too small).
\end{enumerate}

For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.

An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory way, but is learned from the data in an exploratory way. This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypothesis a priori and test them in a confirmatory way, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.

\todo[inline]{\textbf{Previous Papers on IRT and Rasch Models in Journal of School Psychology}

Rasch and IRT models are often mentioned in this journal's papers as being used for scoring school assessments (e.g., Virginia Standards of Learning Assessments;  Oregon Assessment of Knowledge and Skills; Peabody individual achievement test, \cite{Luth92}).

A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}. \edc{included}
There are also papers describing extensions to mixure IRT models \cite{AyalySant17} \edc{included}
and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.

\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200). \edc{included}
}

@@

In mixed-effects models contain two types of effects: fixed effects and random effects. Fixed effects are typically used, for example, to describe the effect of an intervention on the average student performance. Random effects are used to describe the variation in the intercept and slope for, e.g., students in different classes.

When mixed-effects models are combined with MOB, ...

\todo[inline]{Marjolein please edit and finish}

@@

Before being able to use a psychological or educational test for comparing different groups of persons in a fair way, different assumptions of the measurement model have to be checked. In the framework of IRT and Rasch measurement, test items are typically investigated with respect item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}. Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance. Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.

MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model. For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustrated in application example 2. We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model. In the case of the Rasch model, this will be the test items. Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.

If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits. However, in certain settings in educational research, such as large scale assessements, very large sample sizes are available for testing for DIF. Large sample sizes are good for detecting even small effects or model violations with a high statistical power. The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected with large samples. However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough. As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts. Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF. \citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing. In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant). \citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.

@@

\newpage
\section{\edc{Application Example 1}: Subgroup Detection in Mixed-Effects Models}
\label{sec:TutorialMixed}

@@

## Create long dataset
pred_vars <- "program"
HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
                  varying = list(paste0("PPVT_Raw", 2*(43:52)),
                                 paste0("AgeTest_Yr", 2*(43:52))),
                  direction = "long", 
                  v.names = c("PPVT_Raw", "AgeTest_Yr"))
HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
                   "ChildID", "MotherID", "Program", "PPVT", "Age")
HS_dat <- HS_dat[complete.cases(HS_dat), ]
HS_dat$Age <- sqrt(HS_dat$Age)
rownames(HS_dat) <- NULL
saveRDS(HS_dat, file = "HS_dat.Rda")
@

@@

To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start. Head Start is a federally funded nationwide \todo{what nation?} pre-school program for children from low-income families. \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).

The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program. Data from children in those families who participated in another preschool program were excluded. As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start. As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time. As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this? (Mirka: Maybe we can add that the sqrt or the log is a typical way to model trajectories that first increase rapidly, but then later increase more slowly?)}

@@

>>>>

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). %DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness. % Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical model between groups of persons. It generalizes the principle of recursive partitioning, that is also used in classification and regression trees \citep[CART; ][]{Breetal:1984}. 
In CART, the aim is to detect groups of persons, defined by (combinations of) covariate values, that differ in their mean on a response variable. For example, if we want to predict math exam grades using student characteristics assessed in a previous year, we could find that the subgroup of children who are highly motivated and have good reading skills show higher average math exam grades. 
MOB generalizes this idea: It also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean. An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.

The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}. In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}. Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are in turn nested in schools) participate in a study. In application example 1 \ref{sec:TutorialMixed} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure. In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model. We thus assume that the psychological test score(s) of interest \todo{(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)} have already been validated.

For application example 2\todo{add link to section? not sure how to do}, we go back in the research process to the point where a new test has been administered to a validation sample. Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model. Validity assessment requires that we make sure that test results are comparable, for example between children of different genders. If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons. This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability. Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF. In application example 2 we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.

In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}. An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}. Mixture modelling aims at identifying latent classes of persons with different properties. It can also be combined with covariates. A comparison of both approaches is given by \citet{FriStrZei:2014}. \todo{Last 5 sentences seem distracting here. Omit or move to discussion?}

\section{The MOB Algorithm}

\begin{figure}
\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
\begin{subfigure}{.5\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
library("colorspace")
set.seed(3)
x1 <- runif(250, min = 5, max = 11)
x2 <- runif(250, min = 2, max = 8)
x3 <- runif(250, min = 4, max = 10)
error <- rnorm(250, sd = 1.5)
y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
boxplot(toy_data$y, cex = .7)
points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17)
@
\end{subfigure}
\label{fig:global}
\end{figure}

The main rationale of MOB is that one global model may not fit all observations in a dataset equally well. In many studies, additional covariates may be available. It may then be possible to uncover subgroups defined by these covariates, and obtain better-fitting models in each of those subgroups \citep{ZeilyHoth08}.

We illustrate the idea using a very simple toy dataset, comprising \Sexpr{nrow(toy_data)} observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$. To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept. Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle. Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.

To detect possible subgroups with different values for the parameters, MOB cycles iteratively through the following steps:

\begin{enumerate}
\setlength\itemsep{0.25em}
\item The model parameters are first estimated jointly for all persons in the current node, starting with the root node containing the full sample.
\item Structural change in the model parameters is assessed with respect to each available covariate.
\item If there is significant structural change, the observations in the current node are split using the covariate associated with the strongest change.
\item Steps 1--3 are repeated recursively in each resulting node until there is no more significant structural change (or the groups becomes too small).
\end{enumerate}

We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates. The resulting tree is shown in Figure~\ref{fig:tree}. In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3). The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup. In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate. Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible. The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.

\begin{figure}
\caption{Decision tree with group-specific parameters in the terminal nodes.}
\begin{subfigure}{.5\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
library("partykit")
toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
plot(tree, gp = gpar(cex = .7))
@
~
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
plot(tree, gp = gpar(cex = .7), type = "simple")
@
\end{subfigure}
\label{fig:tree}
\end{figure}

%The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates. For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups. The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups. The categorical covariate \texttt{Race}, which was coded in three categories in this example, has been divided into two groups of categories: Black and Hispanic vs.~White. For a binary covariate, \todo{like ... -- add name if is there one in the other example?} 
%on the other hand, there is only one possible cutpoint, namely between the two categories.

%While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items. 
%The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.

%Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting. Variables are selected based on a statistical test for parameter differences, also termed a test for structural change. Structural change is present if, for example, the model parameters systematically differ for children from lower vs.~higher income families. At each node, the algorithm will select the covariate showing the strongest structural change as the next splitting variable. Within this variables, the optimal cutpoint is chosen in a separate step.

%While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees. % Think we can leave this sentence out, I'm not convinced readers will consider that other than binary splits may be implemented.

There are two main differences between MOB and and traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}. The first difference relates to the separation of variable and cutpoint selection. The traditional algorithms performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias. That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. More recent algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper. For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.

The second difference relates to the way MOB stops splitting: It uses a criterion of statistical significance. Once there are no more covariates that show a significant structural change in any node, splitting is halted. In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection. Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected. While the traditional algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (i.e., \textit{pruning}), the MOB framework employs significance tests, but it can also (partly) use effect size measures, as we will see in application example 2 \todo{include link}. This allows for stopping tree growing as soon as no significant differences can be detected anymore. Other stopping criteria are based on the number of persons in the end nodes. These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}

An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner. This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers. Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of covariates or possible partitioning variables and to choose the settings of the MOB algorithm. The current paper aims to provide some guidance.

@@

Mixed-effects models contain two types of effects: Fixed and random effects. Fixed effects are typically used to capture population-averaged effects, while random effects are used to capture inter-individual variation deviating from these fixed effects. In many studies, researchers are specifically interested in testing hypotheses relating to the population-averaged effects, while the random effects are included in the model to properly account for inter-individual variation, and correlations between observations within the same unit.

GLMM trees combine MOB and mixed-effects models and were introduced by \cite{FokkySmit18}. Because researchers' interests commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters. The random-effects parameters can be specified as usual and are assumed constant; they are estimated using all observations in the dataset. The MOB algorithm is thus applied to the fixed-effects parameters only and the detected subgroups will only obtain different estimates for the fixed-effects coefficients. This allows for detecting subgroups with different means on the response variable, but also with differential treatment effects or differential growth over time, to name but a few examples. While the 'standard' MOB trees \cite{ZeilyHoth08} allow for partitioning fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models. Further mathematical and computational details about the GLMM tree algorithm are described in \cite{FokkySmit18} and \citep{FokkyZeil23}.

@@

Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups. In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}. \todo{Omit last sentence or move to discussion? I think better to focus on DIF only here.} Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance. Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.

MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model\todo{possibly confusing: covariates are supplied by user, not searched for. Suggestion: by means of testing whether the parameters of the measurement model exhibit significant instability with respect to (sets of) covariates?}. For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper? (Marjolein: I think software should only be mentioned in the applications.)} The usage of this function will be illustrated in application example 2. We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model. In the case of the Rasch model, this will be the test items. Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.

If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits. However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF. Large sample sizes are good for detecting even small effects or model violations with a high statistical power. The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected with large samples. However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough. As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts. Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF. \citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing. In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant). \citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.

@@

\newpage
\section{Application Example 1: Subgroup Detection in Mixed-Effects Models}
\label{sec:TutorialMixed}

@@

## Create long dataset
pred_vars <- "program"
HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
                  varying = list(paste0("PPVT_Raw", 2*(43:52)),
                                 paste0("AgeTest_Yr", 2*(43:52))),
                  direction = "long", 
                  v.names = c("PPVT_Raw", "AgeTest_Yr"))
HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
                   "ChildID", "MotherID", "Program", "PPVT", "Age")
HS_dat <- HS_dat[complete.cases(HS_dat), ]
HS_dat$Age_orig <- HS_dat$Age
#HS_dat$Age <- sqrt(HS_dat$Age)
rownames(HS_dat) <- NULL
saveRDS(HS_dat, file = "HS_dat.Rda")
@

@@

To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start. Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States. Participation in Head Start takes place from ages 3 through 5. \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).

The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program. Data from children in those families who participated in another preschool program were excluded. As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start. The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF), we will this model PPVT trajectories over time. There were an average number of \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.

%Developmental trajectories rarely if ever show a linear pattern over time; they typically show strong increases early on, which level off over time. We therefore took the square root of the child's age in years, to obtain a pattern of approximately linear increases over time. Appendix A shows PPVT scores as a function of time, both before and after the transformation.

@@

\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage. In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}

\todo{Mention that transformations chosen in growth curve modeling may yield different effects. In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
